Boutiques: a flexible framework for automated application integration in computing platforms

### 2016-11-11 paper.tex
Added related work on CWL

\subsection{Configuration files}

A large number of applications rely on configuration files rather than
command-line options to define their input and output
parameters. Indeed, as the number of parameters increases, command
lines rapidly become long and cumbersome whereas configuration files
allow for better structure and documentation.

@@

Boutiques intentionally does not support the definition of complete
workflows from individual applications. Indeed, we believe that
workflow engines are specific applications that need complex languages to
implement the particular functions that they target. Efforts to
specify common workflow languages include the Common Workflow Language
(CWL)~\cite{cwl} and the Interoperable Workflow Intermediate
Representation (IWIR)~\cite{plankensteiner2011iwir}.

% CWL
The Common Workflow Language (CWL) is the closest related work to
Boutiques, in the sense that it also allows the description of command
lines executed in containers. According to Github, the CWL project
started 6 months before Boutiques. The main conceptual differences
between CWL and Boutiques are the following:
\begin{itemize}
\item CWL has a workflow language, Boutiques does not. CWL also has a
  command line tool description language, but its specification is
  influenced by the fact that command line tools are included in a
  specific workflow languages. In Boutiques, the only available
  workflow construct is sub-tasking (``a task may submit other
  tasks''). This is intentional as we envisage that Boutiques tools
  may be composed through any type of workflow engine and language. For instance, this leads to have multiple output types in CWL.
\item CWL specifies command lines formally whereas in Boutiques they
  are assembled from simple string templates. In short, a command line
  in CWL is specified by an executable and an ordered list of
  parameters. While this representation has the advantage of being
  more formal, for instance the executable can be identified
  automatically, it also removes flexibility by imposing that a tool
  only runs a single command. On the contrary, Boutiques allows any
  type of bash operators to be combined on the command line. This
  level of flexibility is important and greatly reduces the ``shim
  syndrome''. Boutiques' template approach allows for configuration files. 
\item CWL is validated using SALAD, which makes the ecosystem slightly
  more complex. For instance, CWL has ``import'' and ``include''
  directives. Boutiques is validated using plain JSON schema. \todo{ontology layer}.
\item In CWL, the applications need to be instrumented as follows, which is not the case in Boutiques:
\begin{itemize}
\item Applications must write in specific directories.
\item Applications must produce a JSON object describing the output structure.
\end{itemize}
\end{itemize}

\todo{Update from comparison summary on Google doc}

>>>>

\subsection{Configuration files}
\label{sec:configuration-files}
A large number of applications rely on configuration files rather than
command-line options to define their input and output
parameters. Indeed, as the number of parameters increases, command
lines rapidly become long and cumbersome whereas configuration files
allow for better structure and documentation.

@@

\subsection{Common Workflow Language}

The Common Workflow Language is the closest piece of related work as
it supports containers. In particular, the Command Line Tool
Description of CWL overlaps with the \boutiques manifest. According to
Github, CWL started 6 months before Boutiques (September 2014 vs. May
2015). The differences highlighted below were identified from version
1.0 of the CWL Command Line Tool
Description\footnote{\url{http://www.commonwl.org/v1.0/CommandLineTool.html}}.

\subsubsection{Conceptual differences}

The following differences are conceptual in the sense that they may
not be easily addressed by CWL or Boutiques developers.

First, CWL has a workflow language whereas \boutiques does not. In
\boutiques, workflows are integrated as any other applications, except
that they may submit other invocations to enable workflow
parallelism. This fundamental difference has consequences on the
complexity of CWL application descriptions and on the possibility to
reuse existing workflows in \boutiques. The adoption of ontologies in
CWL may also be another consequence (see below).

CWL has a formal command-line definition while \boutiques is more
flexible. CWL specifies command lines using an array of executable and
arguments whereas \boutiques only uses a string template. \boutiques'
template approach may create issues in some cases, but it also allows
developers to add simple operations to an application without having
to write a specific wrapper. For instance, a \boutiques command line
may easily include input decompression using, e.g., the tar command in
addition to the main tool command. On the contrary, CWL applications
may only have a single executable. Importantly, \boutiques' template
system allows supporting configuration files (see
Section~\ref{sec:configuration-files}).

CWL uses ontologies while \boutiques does not. Ontologies allow for
richer definitions but they also have an overhead. The main
consequences are the following:
\begin{itemize}
\item CWL uses a specific framework for
validation, called SALAD (Semantic Annotations for Linked Avro Data)
whereas Boutiques uses plain JSON schema. The main goal of SALAD is to
allow ``working with complex data structures and document formats, such
as schemas, object references, and namespaces''. \boutiques only relies
on the basic types required to describe and validate a command line
syntactically. While the use of SALAD certainly allows for
higher-level validation, and may simplify the composition and
validation of complex workflows, it also introduces a substantial
overhead in the specification, and platforms have to use the validator
provided by CWL. On the contrary, a regular JSON validator can be used in \boutiques.
\item  CWL has a rich set of types whereas \boutiques only has simple
types. Again, this may be seen as a feature or as an overhead
depending on the context. \boutiques tries to limit the complexity of
the specification to facilitate its support by platforms where tools
will be integrated.
\end{itemize}

\subsubsection{Major differences}

The following differences are major but they may be addressed by the
CWL and \boutiques developers as they do not undermine the application
description model.

CWL tools have to write in a specific set of directories called
``designated output directory'', ``designated temporary directory''
and ``system temporary directory''. Tools are informed of the location
of such directories through environment variables. Having to write in
specific directories is problematic because tools have to be modified
to enable that. In \boutiques, the path of output files is defined
using a dedicated property.

CWL types are richer, not only
semantically but also syntactically. For instance, files have
properties for basename, dirname, location, path, checksum, etc.

\boutiques supports various types of containers (Docker, Singularity,
rootfs) while CWL supports only Docker. At the same time, CWL’s
requirements are richer: for instance, they may include RAM and disk
while Boutiques is limited to walltime estimate. CWL has hints, i.e.,
recommendations that only lead to warnings when not respected.

In \boutiques, dependencies can be defined among inputs, e.g., to
specify that an input may be used only when a particular flag is
activated. This is a very useful feature to improve validation, in
particular when tools have a lot of options.

In \boutiques, named
groups of inputs can be defined, which improves the presentation of
long parameter lists for the user and enables the definition of more
constraints within groups (e.g. mutually exclusive inputs).

\subsubsection{Other minor differences}

The \boutiques and CWL schemas have other minor differences that we
summarized in a public
spreadsheet\footnote{\url{https://docs.google.com/spreadsheets/d/1eLhpxzaaMPCIUFmV5Trs-WtaMpBO-c6QDQsmhwP7QIE/edit\#gid=0}}.
